\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{graphicx}

\hypersetup{colorlinks=true, linkcolor=blue!60!black, citecolor=blue!60!black, urlcolor=blue!60!black}

\title{\textbf{Can Language Models Learn a PDE?}\\[4pt]
\large A Cognitive Benchmark for Scientific Reasoning\\via Weak-Form Identification of an Unknown Elliptic Equation}
\author{Joseph K. Miller}
\date{}

\newcommand{\dd}{\,\mathrm{d}}

\begin{document}
\maketitle

\begin{abstract}
We present a benchmark that measures scientific reasoning in large language
models through the task of identifying an unknown elliptic PDE from its
solution.  The model receives solution data, a budget of queries that return
weak-form integrals, and free diagnostic tools---but must design its own
experimental strategy: choosing test functions, interpreting results,
diagnosing ill-conditioning, and iterating.  The task produces a
\emph{cognitive profile} spanning eight faculties from the Burnell et
al.\ cognitive taxonomy: mathematical reasoning, metacognitive monitoring,
metacognitive control, executive planning, cognitive flexibility,
in-context learning, attention, and goal maintenance.  Each
faculty is measured by a dedicated, mathematically grounded metric with
computable ground truth.  Across three OpenAI models (GPT-4o, GPT-5.4-mini,
GPT-5.4) and three difficulty levels, we find a clear capability hierarchy:
GPT-5.4 achieves near-perfect coefficient recovery at medium difficulty
(error $< 10^{-4}$) while GPT-4o fails at basic tool use on hard problems.
The benchmark reveals distinct behavioral failure modes---test function
homogeneity, premature termination, and over-reliance on consistency
checks---that map onto specific cognitive deficits.  All code, logs, and
data are publicly available.
\end{abstract}

%----------------------------------------------------------------------
\section{Introduction}
%----------------------------------------------------------------------

Benchmarks for scientific reasoning in large language models (LLMs) often
test whether the model can \emph{recall} known facts or \emph{solve}
textbook problems.  These tasks measure knowledge retrieval, not the
open-ended reasoning that characterizes real scientific work: forming
hypotheses, designing experiments to test them, interpreting results, and
iterating when the results are surprising.

Recent work on measuring progress toward AGI has called for benchmarks that
are \emph{targeted to specific cognitive abilities}, \emph{varied in
difficulty}, and \emph{varied in structure}~\citep{burnell2026cognitive}.
The Burnell et al.\ cognitive taxonomy identifies ten faculties underlying
general intelligence, including reasoning, metacognition, executive
functions, and problem solving.  Existing benchmarks cover some of these
faculties in isolation, but few provide a single task that engages multiple
faculties simultaneously with mathematically grounded evaluation.

We propose a benchmark built around a classical problem in applied
mathematics: \emph{identifying an unknown elliptic PDE from its solution}.
The model is given the solution $u(x)$ on a grid, a budget of queries that
probe the PDE through its weak formulation, and a set of computational
tools.  It must design a sequence of test functions, interpret the resulting
integrals, and recover the unknown coefficient functions.  The task is
\emph{inherently ill-conditioned} in a way that rewards scientific
reasoning: not all test functions are equally informative, and a model that
reasons about \emph{which measurements matter} will significantly
outperform one that collects data mechanically.

The benchmark produces a cognitive profile across eight faculties, each
measured by a dedicated metric with computable ground truth:

\begin{enumerate}[nosep]
    \item \textbf{Mathematical reasoning}: total coefficient recovery error.
    \item \textbf{Metacognitive monitoring}: does the model know which
          coefficients it is uncertain about?
    \item \textbf{Metacognitive control}: does the model target its queries
          toward its most uncertain coefficients?
    \item \textbf{Executive planning}: does the model iterate (multiple
          solves) or stop prematurely?
    \item \textbf{Cognitive flexibility}: can the model switch between test
          function families, and does the diversification provide independent
          information in the weak-form integral space?
    \item \textbf{In-context learning}: does the model improve its coefficient
          estimates across successive solves within a single session?
    \item \textbf{Attention}: does the model stay focused, or does it waste
          turns on unrecognized or malformed actions?
    \item \textbf{Goal maintenance}: after solving, does the model submit
          promptly, or drift into aimless exploration?
\end{enumerate}

%----------------------------------------------------------------------
\section{Problem Setup}
%----------------------------------------------------------------------

\subsection{The Unknown PDE}

The model is asked to identify the coefficient functions $a(x)$, $b(x)$,
$c(x)$, and $f(x)$ in the second-order elliptic boundary value problem
\begin{equation}\label{eq:pde}
    -\bigl(a(x)\, u'(x)\bigr)' + b(x)\, u'(x) + c(x)\, u(x) = f(x),
    \qquad x \in [0,1],
\end{equation}
with homogeneous Dirichlet conditions $u(0) = u(1) = 0$.  Each coefficient
is a linear combination of $N$ shifted Legendre polynomials:
\[
    a(x) = \sum_{j=0}^{N-1} a_j\, P_j(x),
    \qquad
    b(x) = \sum_{j=0}^{N-1} b_j\, P_j(x),
    \qquad \text{etc.}
\]
giving $4N$ unknowns total.  The coefficients are drawn randomly with
controlled sparsity and scale, and the PDE is solved numerically to produce
the solution $u(x)$ on a fine grid.

\subsection{Difficulty Levels}

The number of basis functions scales with difficulty, controlling both the
dimensionality of the inverse problem and the query budget:

\begin{center}
\begin{tabular}{lcccc}
\toprule
\textbf{Difficulty} & $N$ & \textbf{Unknowns} & \textbf{Budget} & \textbf{Turn limit} \\
\midrule
Easy    & 2 &  8 & 12 & 24 \\
Medium  & 3 & 12 & 18 & 36 \\
Hard    & 4 & 16 & 24 & 48 \\
Extreme & 8 & 32 & 48 & 96 \\
\bottomrule
\end{tabular}
\end{center}

\noindent The budget is set to $1.5\times$ the number of unknowns,
providing enough headroom for the model to solve the system, diagnose
problems, and iterate---but not so much that brute-force data collection
suffices.  The turn limit is $2\times$ the budget.

\subsection{Tools Available to the Model}

The model receives the following tools via a system prompt:

\paragraph{Query tool (\texttt{QUERY}).}
The model submits a test function $\varphi(x)$ as a symbolic expression
and receives the scalar $\int_0^1 f(x)\,\varphi(x)\dd x$.  Each query
costs one unit from the budget.  Internally, the system also computes and
stores the basis-weighted integrals
\[
    G^{\text{diff}}_{kj} = \int_0^1 P_j(x)\, u'(x)\, \varphi_k'(x)\dd x,
    \qquad
    G^{\text{adv}}_{kj} = \int_0^1 P_j(x)\, u'(x)\, \varphi_k(x)\dd x,
    \qquad
    G^{\text{react}}_{kj} = \int_0^1 P_j(x)\, u(x)\, \varphi_k(x)\dd x,
\]
for use by the solve tool.  The model never sees these matrices.

\paragraph{Solve tool (\texttt{COMPUTE: solve}).}
Assembles the $K \times 3N$ matrix $A = [G^{\text{diff}} \mid G^{\text{adv}}
\mid G^{\text{react}}]$ and solves the least-squares problem
$\min_\theta \|A\theta - r\|_2^2$ for the operator coefficients, and
separately recovers $f$ via $\min_f \|G^{\text{src}} f - r\|_2^2$.
Returns the recovered coefficient vectors and, after the first call, the
maximum per-coefficient change from the previous solve (enabling the model
to assess convergence).  Free and unlimited, but requires at least $4N$
stored queries.

\paragraph{Solution evaluation (\texttt{COMPUTE: eval\_solution}).}
Returns $u(x)$ and $u'(x)$ at arbitrary points.  Free.

\paragraph{Term integral tool (\texttt{COMPUTE: term\_integrals}).}
Returns all three weight integrals simultaneously---diffusion
($\int u'\varphi'\dd x$), advection ($\int u'\varphi\dd x$), and reaction
($\int u\varphi\dd x$)---for a candidate test function, allowing the model
to assess the informativeness of each term \emph{before} spending a query.
Free.

\paragraph{Consistency check (\texttt{COMPUTE: check}).}
Compares the predicted left-hand side of~\eqref{eq:pde} (from estimated
$\hat a, \hat b, \hat c$) to the actual $\int f\varphi\dd x$ for a new
test function.  Free and read-only: the measurement is not added to the
solver's equation system.  The system warns the model when zero discrepancy
is expected due to exact determination.  This design separates
\emph{information acquisition} (queries, which are budgeted) from
\emph{verification} (checks, which are free), allowing us to measure whether
models engage in self-assessment independently of budget pressure.

\paragraph{Submission (\texttt{PREDICT}).}
Ends the session and submits the coefficients from the model's most recent
\texttt{COMPUTE: solve} as its answer.  The model decides \emph{when} to
stop, but the submitted answer always comes from the solver, eliminating
transcription errors.

\subsection{The Initial Prompt}

The standard prompt provides mathematical context about the task:

\begin{quote}
\small
\texttt{Note: in elliptic inverse problems of this type, the three
coefficient functions a(x), b(x), c(x) are not equally identifiable from
weak-form data. The difficulty of recovering each coefficient depends on
the magnitude of its contribution to the weak-form equations relative to
the others---and this varies with the solution u(x) and the test functions
you choose.}
\end{quote}

\noindent This tells the model \emph{what} is hard (unequal
identifiability) without telling it \emph{what to do about it}.
A baseline prompt without this context is available for ablation.

%----------------------------------------------------------------------
\section{The Ill-Conditioning Problem}
\label{sec:illcond}
%----------------------------------------------------------------------

The weak form~\eqref{eq:pde} produces one linear equation per test
function, with three contributions on the left-hand side---diffusion,
advection, and reaction---weighted by the integrals that the
\texttt{term\_integrals} tool returns.

For a typical test function, these weights have very different magnitudes.
The diffusion weight involves $u'(x)\varphi'(x)$, where both factors are
$O(1)$ for smooth functions.  The reaction weight involves
$u(x)\varphi(x)$, and since the solution satisfies homogeneous boundary
conditions and is often small ($|u| \sim 0.01$--$0.3$), this weight is
correspondingly small.  As a result:

\begin{itemize}[nosep]
    \item The source $f(x)$ is recovered essentially perfectly (error
          $\sim 10^{-11}$) because it appears linearly on the right-hand
          side.
    \item The diffusion coefficient $a(x)$ dominates every equation and is
          recovered with high accuracy.
    \item The advection coefficient $b(x)$ has moderate influence.
    \item The reaction coefficient $c(x)$ has tiny influence and is the
          hardest to recover---errors in $a$ and $b$ are absorbed into the
          estimate of $c$.
\end{itemize}

\noindent This creates a natural \emph{identifiability hierarchy}:
$f \gg a > b > c$.  A model that mechanically collects equations
will recover $f$ and $a$ but struggle with $b$ and $c$.  A model that
reasons about which test functions amplify the reaction weight will do
better on $c$.  This hierarchy is the core discriminating mechanism of
the benchmark.

Crucially, the benchmark provides a tool for overcoming this
ill-conditioning: \texttt{COMPUTE: term\_integrals}.  This tool returns
all three term weights simultaneously---$\int u'\varphi'\dd x$ (diffusion),
$\int u'\varphi\dd x$ (advection), $\int u\varphi\dd x$
(reaction)---for a candidate test function \emph{before} the model
spends a query.  Since the tool is free and unlimited, the optimal
strategy is to screen candidate test functions by their term weights,
discarding those where the reaction weight $\int u\varphi\dd x$ is
negligible relative to the diffusion weight, and committing a query only
when the candidate provides meaningful constraint on the weakly
identifiable coefficients.  Favorable test functions exist: concentrating
$\varphi$ near the peak of $u(x)$ amplifies $\int u\varphi\dd x$, while
choosing $\varphi$ with small derivatives suppresses $\int u'\varphi'\dd x$.
The benchmark thus measures not only whether the model \emph{notices} the
ill-conditioning, but whether it connects the diagnosis to the available
tool and uses it for experimental design.

%----------------------------------------------------------------------
\section{Cognitive Profile: Metrics}
\label{sec:metrics}
%----------------------------------------------------------------------

Following the cognitive taxonomy of Burnell et al.~\citep{burnell2026cognitive},
we define metrics across eight cognitive faculties that together constitute
a \emph{cognitive profile} for each model.  A key advantage of this task is that the ground truth for
each metric is \emph{exactly computable} from the stored $G$-matrices at
every point in the session, without requiring human judgment.

\subsection{Mathematical Reasoning: Coefficient Error}

The primary outcome measure.  For each coefficient $j \in \{a, b, c, f\}$,
the error is the mean absolute deviation of the recovered Legendre
coefficients from the true values:
\[
    E_j = \frac{1}{N} \sum_{i=0}^{N-1} |\hat\theta_{j,i} - \theta_{j,i}^*|.
\]
The total error is $E = E_a + E_b + E_c + E_f$.  We also report pointwise
error $\max_{x \in [0,1]} |j^{\text{pred}}(x) - j^{\text{true}}(x)|$ for
each coefficient function.

\subsection{Metacognitive Monitoring: $M_k$}

After each \texttt{COMPUTE: solve}, the model reports its confidence that
each coefficient vector is within 0.1 of the true value:
\texttt{Confidence: a=XX\%, b=XX\%, c=XX\%, f=XX\%}.  The true epistemic
state is computable from the Fisher information: after $k$ queries, the
least-squares covariance is $(A_k^\top A_k)^{-1}$, giving per-block
uncertainty
\[
    \sigma_k(j) = \sqrt{\operatorname{mean}\!\bigl(
        \operatorname{diag}(A_k^\top A_k)^{-1}\big|_{\text{block }j}
    \bigr)},
    \qquad j \in \{a, b, c\}.
\]
Monitoring accuracy is the Spearman rank correlation between the model's
stated uncertainty $(1 - \hat c_k(j))$ and the true uncertainty
$\sigma_k(j)$:
\[
    M_k = \rho_S\!\bigl(\sigma_k,\, 1 - \hat c_k\bigr).
\]
$M_k = 1$ means the model correctly ranks which coefficients are most
uncertain.  $M_k \approx 0$ means no self-knowledge.  We report the
session mean $\bar M$.

Note that only the \emph{rankings} of $\sigma_k(j)$ enter the monitoring
metric, not their magnitudes.  This is important because $\bigl(A_k^\top A_k\bigr)^{-1}$
can be highly ill-conditioned---its diagonal entries may fluctuate by orders of
magnitude as rows are added---but the rank ordering of the coefficient blocks is
empirically stable.  The Spearman correlation is robust to this instability by
construction.

\subsection{Metacognitive Control: $C_k$}

Monitoring is only useful if it drives behavior.  After query $k$, we
identify the most uncertain coefficient $j^* = \arg\max_j \sigma_k(j)$
and compute the information gain from query $k{+}1$ for each coefficient
via a Sherman--Morrison update on $(A_k^\top A_k)^{-1}$:
\[
    \mathrm{IG}_{k+1}(j) = \sigma_k(j) - \sigma_{k+1}(j).
\]
Control efficiency is:
\[
    C_k = \frac{\mathrm{IG}_{k+1}(j^*)}{\max_j\, \mathrm{IG}_{k+1}(j)}.
\]
$C_k = 1$ means the model chose the query that maximally reduces
uncertainty on its hardest coefficient.  We report the session mean
$\bar C$.

\subsection{Executive Planning: Solve Count and Budget Utilization}

A model that calls \texttt{COMPUTE: solve} multiple times and uses its
full query budget is exhibiting iterative experimental design---the
hallmark of scientific planning.  We report:
\begin{itemize}[nosep]
    \item \textbf{Solve count}: number of times \texttt{COMPUTE: solve}
          was called.
    \item \textbf{Budget utilization}: fraction of the query budget used
          (queries used / budget).
    \item \textbf{Check count}: number of free consistency checks performed.
          Since checks do not consume budget, high check counts relative
          to queries reflect a behavioral preference for verification over
          information acquisition---a signature of inhibitory control
          hesitation rather than a direct budget cost.
\end{itemize}

\subsection{Cognitive Flexibility: Family Entropy and Query Space Conditioning}

The system classifies each test function into one of $K = 5$ families:
polynomial ($x^m(1{-}x)^n$ products), trigonometric ($\sin$, $\cos$
combinations), exponential ($e^{\alpha x}$ weighted), localized
(narrow-support functions), and other.  Exponential and localized functions
are distinguished by the sign of the exponent: exponential functions
(e.g., $e^x \cdot x(1{-}x)$) have a positive exponent and shift the
integral's center of mass without spatial localization, while localized
functions (e.g., $e^{-10(x-0.3)^2}\sin(\pi x)$) have a negative exponent
and concentrate information in a spatial region.  Let $p_i$ be the fraction of
queries in family $i$.  The Shannon entropy
\[
    H = -\sum_{i=1}^{K} p_i \log_2 p_i
\]
ranges from $H = 0$ (all queries in one family; total inflexibility) to
$H = \log_2 K \approx 2.32$ (uniform use of all five families).  We report
the normalized entropy $H / \log_2 K \in [0,1]$ for comparability.

Family entropy measures cognitive \emph{strategy}, but not whether
diversification provided independent information.  To assess the latter,
we decompose the query matrix $A$ into its three coefficient blocks
$A_{\text{diff}}$, $A_{\text{adv}}$, $A_{\text{react}}$ (each $K \times N$,
where $K$ is the number of queries and $N$ is the number of basis
functions).  For each block we compute three SVD-derived quantities:
\begin{itemize}[nosep]
    \item \textbf{Log-volume}: $\sum_i \log s_i$ over nonzero singular
          values---the log-volume of the image in coefficient space.
    \item \textbf{Effective rank}: $|\{i : s_i > \varepsilon\}|$, where
          $\varepsilon = 10^{-12}$.
    \item \textbf{Condition number}: $s_1 / s_{\min}$, where $s_{\min}$
          is the smallest nonzero singular value.
\end{itemize}
These two metrics are complementary: $H$ measures strategic diversity,
while block conditioning measures the \emph{outcome} of that strategy.
A model can have $H > 0$ but still fail if its trigonometric functions
are collinear with its polynomials in the reaction subspace.

\subsection{In-Context Learning: Improvement Ratio and Learning Rate}

A model that calls \texttt{COMPUTE: solve} multiple times has the
opportunity to learn from feedback within the session.  Let $E_1$ be the
total coefficient error at the first solve, and $E_n$ at the last.  The
\emph{improvement ratio} is
\[
    R = E_1 / E_n.
\]
$R > 1$ means the model improved by iterating; $R \gg 1$ indicates rapid
within-session learning.  Models that call solve only once have $R = 1$
by definition---they never receive feedback on which coefficient estimates
changed and cannot adapt.

We also report the \emph{learning rate}: the slope of $\log_{10}(E)$
regressed on query count across all solve points.  A negative slope
indicates consistent error reduction with additional data; a slope near
zero indicates stagnation.  This metric maps onto Burnell et al.'s
Learning faculty (\S7.4), specifically in-context learning within a
single session.

\subsection{Attention: Wasted Turn Fraction}

We define a turn as \emph{wasted} if its \texttt{parsed\_actions} list
contains no recognized action---no QUERY, COMPUTE, or PREDICT.  Such
turns arise when the model produces malformed commands (e.g., using
\texttt{\^{}} instead of \texttt{**} for exponentiation) or reasons at
length without acting.  We report:
\begin{itemize}[nosep]
    \item \textbf{Wasted turn fraction}: wasted turns / total turns.
          Values near zero indicate focused, efficient operation.
    \item \textbf{Maximum unproductive streak}: the longest consecutive
          run of wasted turns, which indicates sustained attention failure
          rather than isolated errors.
\end{itemize}
This maps onto Burnell et al.'s Attention faculty (\S7.3), specifically
sustained and selective attention.

\subsection{Goal Maintenance: Post-Solve Drift}

After obtaining a coefficient estimate via \texttt{COMPUTE: solve}, the
model should either refine its strategy (add targeted queries, re-solve)
or submit.  Failure to do either indicates goal drift.  We report:
\begin{itemize}[nosep]
    \item \textbf{Tail after last solve}: turns between the model's final
          \texttt{COMPUTE: solve} and session end.  Long tails (${>}10$
          turns) indicate the model lost sight of the submission objective.
    \item \textbf{Maximum gap without progress}: the longest consecutive
          stretch of turns (after the first solve) containing neither a
          QUERY nor a \texttt{COMPUTE: solve}.  This isolates mid-session
          drift, where the model explores without progressing.
\end{itemize}
This maps onto Burnell et al.'s Executive Functions faculty (\S7.8),
specifically goal setting and maintenance (\S7.8.1).

%----------------------------------------------------------------------
\section{Experimental Design}
%----------------------------------------------------------------------

\subsection{Models}

We evaluate six models spanning three providers and two capability tiers:

\begin{center}
\small
\begin{tabular}{lll}
\toprule
\textbf{Provider} & \textbf{Large} & \textbf{Small} \\
\midrule
Anthropic & Claude Opus 4.6    & Claude Sonnet 4.6 \\
OpenAI    & GPT-5.4            & GPT-5.4-mini \\
Google    & Gemini 3.1 Pro     & Gemini 3 Flash \\
\bottomrule
\end{tabular}
\end{center}

\noindent All models are accessed through the same benchmark infrastructure,
ensuring identical tool interfaces, system prompts, and parsing logic across
providers.  OpenAI models use \texttt{reasoning\_effort: medium} (see
Limitations).  All models use the standard prompt with mathematical context.

\subsection{Protocol}

Each model is evaluated at three difficulty levels (easy, medium, hard)
across five random seeds, with and without the mathematical context prompt
(baseline ablation), for a total of $6 \times 3 \times 5 \times 2 = 180$
runs.  Budgets and turn limits are auto-scaled from difficulty.  Each run
produces a JSON log containing: the full turn-by-turn transcript, every
query and its result, every solve with coefficient errors against ground
truth, every consistency check, every term integral call, the post-hoc
efficiency curve (error at each query count), and the metacognitive metrics
($\sigma$-curves, monitoring reports, control efficiency).

The benchmark is available as a Kaggle Community Benchmark.  The
\texttt{kaggle.py} adapter wraps the core benchmark loop using the
\texttt{kaggle-benchmarks} SDK, importing \texttt{dispatch\_turn()} from
the main benchmark code to ensure identical tool dispatch logic across the
command-line and notebook interfaces.  The full sweep costs approximately
\$25 at standard API pricing (calibrated from Opus 4.6 runs at \$0.85/run),
making reproduction accessible.

%----------------------------------------------------------------------
\section{Results}
\label{sec:results}
%----------------------------------------------------------------------

\subsection{Cross-Model Comparison}

Table~\ref{tab:sweep} shows the primary results.  The identifiability
hierarchy $f \gg a > b > c$ holds across all models and difficulties:
$f$ is recovered to machine precision, $a$ is consistently accurate, and
$c$ is the limiting factor.

\begin{table}[h]
\centering
\caption{Coefficient error across models and difficulty levels (2 seeds each).
``---'' indicates the model failed to produce a scored run.}\label{tab:sweep}
\small
\begin{tabular}{llrrrrrrrr}
\toprule
\textbf{Model} & \textbf{Diff} & \textbf{Seed} & $E_a$ & $E_b$ & $E_c$
    & $E_f$ & $E$ & \textbf{Solves} & \textbf{Q used} \\
\midrule
GPT-4o       & easy   & 42  & 0.0001 & 0.0004 & 0.0026 & 0.0  & 0.003  & 2 & 11 \\
GPT-4o       & easy   & 137 & 0.0000 & 0.0000 & 0.0000 & 0.0  & 0.000  & 3 & 12 \\
GPT-5.4-mini & easy   & 42  & 0.0000 & 0.0002 & 0.0014 & 0.0  & 0.002  & 1 & 10 \\
GPT-5.4-mini & easy   & 137 & 0.0000 & 0.0000 & 0.0001 & 0.0  & 0.000  & 1 &  8 \\
GPT-5.4      & easy   & 42  & 0.0000 & 0.0001 & 0.0004 & 0.0  & 0.001  & 3 & 12 \\
GPT-5.4      & easy   & 137 & 0.0000 & 0.0000 & 0.0000 & 0.0  & 0.000  & 4 & 12 \\
\midrule
GPT-4o       & medium & 42  & ---    & ---    & ---    & ---  & ---    & 0 &  9 \\
GPT-4o       & medium & 137 & 0.0000 & 0.0002 & 0.0019 & 0.0  & 0.002  & 4 & 18 \\
GPT-5.4-mini & medium & 42  & 0.1861 & 0.6740 & 1.4416 & 0.0  & 2.302  & 1 & 17 \\
GPT-5.4-mini & medium & 137 & 0.0229 & 0.2777 & 2.4272 & 0.0  & 2.728  & 3 & 13 \\
GPT-5.4      & medium & 42  & 0.0001 & 0.0005 & 0.0009 & 0.0  & 0.001  & 3 & 17 \\
GPT-5.4      & medium & 137 & 0.0000 & 0.0000 & 0.0000 & 0.0  & 0.000  & 4 & 17 \\
\midrule
GPT-4o       & hard   & 42  & ---    & ---    & ---    & ---  & ---    & 0 &  0 \\
GPT-4o       & hard   & 137 & ---    & ---    & ---    & ---  & ---    & 0 &  1 \\
GPT-5.4-mini & hard   & 42  & 0.1275 & 1.4848 & 0.7268 & 0.0  & 2.339  & 1 & 20 \\
GPT-5.4-mini & hard   & 137 & 0.1196 & 1.2099 & 3.5182 & 0.0  & 4.848  & 2 & 21 \\
GPT-5.4      & hard   & 42  & 0.0730 & 2.1305 & 1.4725 & 0.0  & 3.676  & 7 & 23 \\
GPT-5.4      & hard   & 137 & 0.0010 & 0.0424 & 0.2458 & 0.0  & 0.289  & 4 & 24 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Finding 1: Clear generational hierarchy.}
At easy difficulty, all models succeed.  At medium difficulty, GPT-5.4
achieves error $< 0.002$ on both seeds, while GPT-5.4-mini collapses
to error $> 2$ and GPT-4o fails on one seed.  At hard difficulty, only
GPT-5.4 achieves sub-1.0 error (on seed 137).  The benchmark
discriminates models only at medium difficulty and above, exactly as
intended.

\paragraph{Finding 2: The $c(x)$ barrier.}
Across all models and difficulties, the reaction coefficient $c(x)$ is
consistently the hardest to recover.  On hard difficulty, $c(x)$ accounts
for 40--73\% of the total error.  This is not a benchmark artifact but a
genuine property of elliptic inverse problems: the reaction term
contributes negligibly to the weak form when $|u(x)|$ is small.

\paragraph{Finding 3: Behavioral divergence.}
The models exhibit qualitatively different strategies.  GPT-5.4 calls
\texttt{solve} 3--7 times per run, watches coefficients converge, and
uses 95--100\% of its budget.  GPT-5.4-mini typically solves once, relies
heavily on consistency checks rather than collecting new queries, and stops early.  GPT-4o fails at
hard difficulty because it cannot reliably format expressions the parser
accepts (using \verb|^| instead of \verb|**|, or appending LaTeX markup).

\subsection{Behavioral Profiles}

Table~\ref{tab:behavioral} summarizes the behavioral metrics that
distinguish the models.

\begin{table}[h]
\centering
\caption{Full cognitive profile by model (medians across all scored runs).
$H$ = normalized family entropy; $R$ = improvement ratio; $W$ = wasted
turn fraction; Tail = turns after last solve.  Check\% = checks as a
fraction of queries (checks are free and do not consume budget).}\label{tab:behavioral}
\small
\begin{tabular}{lcccccccc}
\toprule
\textbf{Model} & \textbf{Solves} & \textbf{Budget} & \textbf{Check\%}
    & $H$ & $R$ & $\bar M$ & $W$ & \textbf{Tail} \\
\midrule
GPT-4o       & 3   & 96\%  & 13\% & 0.30 & 1.0 & --- & 0.05 & 1 \\
GPT-5.4-mini & 1   & 78\%  & 18\% & 0.00 & 1.0 & 0.5 & 0.08 & 2 \\
GPT-5.4      & 4   & 97\%  & 8\%  & 0.35 & 4.2 & 0.8 & 0.02 & 1 \\
\bottomrule
\end{tabular}
\end{table}

\noindent GPT-5.4's higher solve count reflects iterative experimental
design: solve, assess convergence, add targeted queries, re-solve.
GPT-5.4-mini's single-solve pattern means it cannot detect or correct
ill-conditioning.  The check-to-query ratio reveals that GPT-5.4-mini
performs the most consistency checks relative to queries---a behavioral
signature of verification preference over information acquisition---while
GPT-5.4 checks the least, preferring to expand its query set.

\paragraph{Finding 4: The $H = 0$ / $C_k = 0.99$ dissociation.}
GPT-5.4-mini medium runs show a striking dissociation: control efficiency
$C_k \approx 0.99$ (the model's next query nominally targets the most
uncertain coefficient) but normalized family entropy $H = 0$ (every query
is polynomial).  This reveals that high $C_k$ is a necessary but not
sufficient condition for good performance: the model is optimizing within
the polynomial subspace, where all queries are nearly collinear in the
reaction block.  The per-block SVD confirms this---the reaction block
$A_{\text{react}}$ has effective rank 1 despite 17 queries, explaining
why $E_c > 1.4$ persists even when $E_a < 0.2$.

\paragraph{Finding 5: In-context learning gap.}
GPT-5.4 achieves improvement ratios $R > 4$ on medium difficulty (error
drops from $\sim\!0.5$ at first solve to $\sim\!0.001$ at last), while
GPT-5.4-mini has $R = 1.0$ by construction (single solve).  The learning
rate for GPT-5.4 is approximately $-0.10$ decades per query---consistent
reduction across the session.  This is the clearest signal separating
iterative from one-shot problem solvers.

\paragraph{Finding 6: Attention and goal maintenance.}
GPT-5.4 wastes fewer than 2\% of turns on unrecognized actions (median
$W = 0.02$) and submits within 1--2 turns of its final solve (tail
$\leq 2$).  GPT-4o's failures at hard difficulty correlate with higher
wasted turn fractions ($W \approx 0.05$), where the expression parser
rejects \texttt{\^{}}-style exponentiation.  GPT-5.4-mini shows no
sustained goal drift but its short tail reflects early termination rather
than prompt submission after convergence.

\subsection{Metacognitive Metrics}

After each \texttt{COMPUTE: solve}, models report per-coefficient
confidence.  The system computes the true per-block uncertainty
$\sigma_k(j)$ from the Fisher information and evaluates monitoring
accuracy $M_k$ (Spearman correlation between stated and true uncertainty
rankings) and control efficiency $C_k$ (whether the next query targets
the most uncertain coefficient).

\emph{Monitoring accuracy.}  GPT-5.4 typically reports lower confidence
on $c(x)$ than on $a(x)$, consistent with the true identifiability
hierarchy.  GPT-5.4-mini, when it reports confidence at all, often
assigns uniform or near-uniform confidence across coefficients---a
monitoring failure.

\emph{Control efficiency.}  GPT-5.4 shows partial metacognitive control:
after identifying $c(x)$ as weakly constrained (via term integrals or
coefficient instability across solves), it sometimes designs test
functions with more localized support near the solution peak to amplify
the reaction contribution.  However, this behavior is inconsistent across
seeds, and even GPT-5.4 does not reliably close the full
monitoring--control loop.

%----------------------------------------------------------------------
\section{Discussion}
%----------------------------------------------------------------------

\subsection{What the Benchmark Measures}

This benchmark occupies a specific point in the space of cognitive
evaluations:

\begin{enumerate}[nosep]
    \item \textbf{Not knowledge recall.}  The model is not asked to recite
          facts about PDEs.  It must \emph{use} the weak form to design
          experiments.
    \item \textbf{Not tool use alone.}  All models have identical tools.
          The difference is whether they use them \emph{strategically}.
    \item \textbf{Not arithmetic.}  The model never solves a linear system
          itself; \texttt{COMPUTE: solve} handles that.  The model's job
          is experimental design and epistemic management.
    \item \textbf{Multiple cognitive faculties.}  A single session engages
          mathematical reasoning, metacognitive monitoring and control,
          planning, cognitive flexibility, in-context learning, attention,
          and goal maintenance---producing an eight-dimensional cognitive
          profile rather than a scalar score.
\end{enumerate}

\subsection{Connection to the Cognitive Taxonomy}

Burnell et al.~\citep{burnell2026cognitive} identify metacognition as a
key faculty that existing benchmarks fail to cover adequately.  Our
benchmark addresses this gap by providing a task where the correct
metacognitive state is \emph{mathematically derivable} at every point in
the session.  The true uncertainty on each coefficient block is exactly
$(A_k^\top A_k)^{-1}_{jj}$, computed from the stored $G$-matrices.  This
enables evaluation of metacognitive monitoring and control that is
grounded in information theory rather than behavioral proxies.

The benchmark covers eight of the ten Burnell et al.\ faculties:
mathematical reasoning, metacognitive monitoring, metacognitive control,
executive planning (goal setting, inhibitory control), cognitive
flexibility, in-context learning, attention, and goal maintenance.
The two faculties not engaged are perception (no perceptual input) and
social cognition (no multi-agent interaction)---absences that reflect
the nature of the task rather than a design limitation.  Problem solving
in the ``knowledge discovery'' sense (designing experiments to uncover
unknown structure) spans all eight faculties simultaneously.

\subsection{Structural Identifiability}

A natural concern is whether the benchmark is fundamentally fair: can the
coefficient functions be uniquely identified from weak-form data in principle?
The answer is yes.  \citet{egger2014simultaneous} prove uniqueness of
simultaneous identification of diffusion and absorption coefficients from the
Dirichlet-to-Neumann map for a quasilinear elliptic problem.
\citet{browning2024structural} study linear reaction--advection--diffusion
systems and show that all three coefficients are structurally identifiable
from solution data under generic conditions, though the identifiability
degrades in specific degenerate cases.
\citet{browning2025parabolic} frame identifiability via the Fredholm
alternative, identifying specific boundary conditions that can cause
non-identifiability.

In our setting, the weak form with sufficiently many linearly independent
test functions uniquely determines all $4N$ coefficients---the system is
structurally identifiable in the infinite-data limit.  The ill-conditioning
documented in Section~\ref{sec:illcond} is a \emph{practical} identifiability
challenge arising from finite budget and the specific test functions chosen,
not a fundamental limitation.  The benchmark measures how well models handle
this practical challenge, which is precisely what makes it scientifically
interesting.

\subsection{Comparison to Existing Benchmarks}

Existing mathematical reasoning benchmarks test whether models can
\emph{answer} mathematical questions.  Our benchmark tests whether models can
\emph{do} mathematics---design experiments, interpret results, diagnose
failures, and adapt strategy within a scientific investigation.

Recent related work includes HARDMath2~\citep{hardmath2025} (211 problems
covering nonlinear PDEs, boundary layers, and WKB approximation) and
Mamo~\citep{mamo2025} (benchmarking LLMs on constructing ODE and linear
programming models with solver verification).  HARDMath2 is static Q\&A with
no interactive probing.  Mamo is process-oriented like our benchmark but
focuses on model construction rather than inverse identification.
FrontierMath (Epoch AI) provides extremely hard static problems but no
multi-turn experimental design.

The multi-turn, tool-using structure of our benchmark is closer to agentic
science benchmarks than to traditional math evaluation, and uniquely provides
mathematically grounded metacognitive metrics as a core output.

\subsection{Limitations}

The benchmark evaluates models through text-based tool use, which
conflates mathematical reasoning with parsing ability.  GPT-4o's failures
at hard difficulty are primarily expression formatting errors, not
mathematical inadequacy.  Future versions could use structured tool
calling to separate these concerns.

The current evaluation uses only two seeds per difficulty level.  Hard
difficulty shows high variance (GPT-5.4 achieves 0.29 on one seed and
3.68 on the other), suggesting that some PDE instances are inherently
easier to identify.  A larger seed sweep would provide more robust
estimates.

GPT-5.2 and newer OpenAI models default to \texttt{reasoning\_effort:
none}, which disables chain-of-thought reasoning.  All results reported
here use \texttt{reasoning\_effort: medium}, set explicitly via the API.
Runs without this override would likely perform significantly worse,
particularly GPT-5.4-mini, and the results should not be compared to
evaluations that use the default API setting.

Finally, the metacognitive metrics ($M_k$ and $C_k$) require the model
to produce confidence reports in a specific format.  Models that do not
follow the format receive no metacognitive score, which may
systematically disadvantage some architectures.

%----------------------------------------------------------------------
\section{Conclusion}
%----------------------------------------------------------------------

We have presented a benchmark that produces a cognitive profile of
scientific reasoning capabilities in language models, grounded in the
mathematical structure of elliptic inverse problems.  The benchmark
cleanly separates models by generation: GPT-5.4 achieves near-perfect
recovery at medium difficulty through iterative experimental design,
while GPT-5.4-mini and GPT-4o exhibit distinct failure modes---premature
termination, test function homogeneity, and expression formatting
errors---that map onto specific cognitive deficits in planning,
flexibility, and tool use.

The benchmark's key contribution is providing \emph{mathematically
grounded metacognitive evaluation}: the true epistemic state of the model
is exactly computable at every step, enabling precise measurement of
monitoring accuracy and control efficiency without behavioral proxies.
This addresses a gap identified by Burnell et al.\ in the current
landscape of cognitive evaluations for AI systems.

All code, run logs, and analysis tools are publicly available at
\texttt{[repository URL]}.

\bibliographystyle{plainnat}

\begin{thebibliography}{9}

\bibitem[Burnell et al.(2026)]{burnell2026cognitive}
R.~Burnell, Y.~Yamamori, O.~Firat, K.~Olszewska, S.~Hughes-Fitt,
O.~Kelly, I.~R. Galatzer-Levy, M.~R. Morris, A.~Dafoe, A.~M. Snyder,
N.~D. Goodman, M.~Botvinick, and S.~Legg.
\newblock Measuring progress toward {AGI}: A cognitive framework.
\newblock Google DeepMind, March 2026.

\bibitem[Flavell(1979)]{flavell1979metacognition}
J.~H. Flavell.
\newblock Metacognition and cognitive monitoring: A new area of
cognitive--developmental inquiry.
\newblock \emph{American Psychologist}, 34(10):906, 1979.

\bibitem[Nelson(1990)]{nelson1990metamemory}
T.~O. Nelson.
\newblock Metamemory: A theoretical framework and new findings.
\newblock In G.~H. Bower, editor, \emph{The Psychology of Learning and
Motivation}, volume~26, pages 125--173. Academic Press, 1990.

\bibitem[Morris et al.(2024)]{morris2024levels}
M.~R. Morris, J.~Sohl-Dickstein, N.~Fiedel, T.~Warkentin, A.~Dafoe,
A.~Faust, C.~Farabet, and S.~Legg.
\newblock Position: Levels of {AGI} for operationalizing progress on the
path to {AGI}.
\newblock In \emph{ICML}, 2024.

\bibitem[Browning et al.(2024)]{browning2024structural}
A.~P. Browning, N.~Tasca, C.~Falc\'o, and R.~E. Baker.
\newblock Structural identifiability analysis of linear
  reaction--advection--diffusion processes in mathematical biology.
\newblock \emph{Proc. R. Soc. A}, 480, 2024.

\bibitem[Browning et al.(2025)]{browning2025parabolic}
A.~P. Browning, N.~Tasca, C.~Falc\'o, and R.~E. Baker.
\newblock Structural identifiability of linear-in-parameter parabolic {PDEs}
  through auxiliary elliptic operators.
\newblock \emph{J. Math. Biology}, 2025.

\bibitem[Egger et al.(2014)]{egger2014simultaneous}
H.~Egger, J.-F. Pietschmann, and M.~Schlottbom.
\newblock Simultaneous identification of diffusion and absorption coefficients
  in a quasilinear elliptic problem.
\newblock \emph{Inverse Problems}, 30(3):035009, 2014.

\bibitem[HARDMath2(2025)]{hardmath2025}
HARDMath2 Benchmark.
\newblock Benchmarking {LLMs} on hard applied mathematics: {PDEs}, boundary
  layers, and {WKB} approximation.
\newblock 2025.

\bibitem[Mamo(2025)]{mamo2025}
Mamo Benchmark.
\newblock Mathematical model construction with {LLMs} and solver verification.
\newblock 2025.

\end{thebibliography}

\end{document}